Young's Experiment

### 2020-11-11 fdtd/em.tex
Final draft

Next we set up the physical parameters.
\begin{verbatim}
ddx = 0.01        # Cell size.
c = 3e8           # Speed of light in vacuum.
dt = ddx / 2*c    # Time step size.
epsz = 8.854e-12  # Permittivity of free space.
t0 = 20           # Mean magnitude of the pulse.
spread = 6        # Spread of the pulse.
\end{verbatim}

@@

plt.show()
\end{verbatim}
The results of the simulation are shown in figure \ref{f3}.
\begin{figure}
\centering
\includegraphics[scale=0.5]{fdtd-2}
\caption{Propagation of a Gaussian pulse in 2d}\label{f3}
\end{figure}
\bibliographystyle{plain}
\bibliography{em}
\end{document}

>>>>

Next we set up the physical parameters.
\begin{verbatim}
ddx = 0.01        # Cell size.
c = 3e8           # Speed of light in vacuum.
dt = ddx / 2*c    # Time step size.
t0 = 20           # Mean magnitude of the pulse.
spread = 6        # Spread of the pulse.
\end{verbatim}

@@

plt.show()
\end{verbatim}
The results of the simulation are shown in figure \ref{f3}.
\begin{figure}
\centering
\includegraphics[scale=0.5]{fdtd-2}
\caption{Propagation of a Gaussian pulse in 2d}\label{f3}
\end{figure}

\section{Simulation of Young's experiment}\label{s9}
The code snippets in this section show how to simulate Young's experiment
using FDTD. This one is also a minor modification of the code in Houle in 
Sullivan's book \cite{sullivan2013electromagnetic}. We start with importing
all packages needed for the simulation. The first three are needed for the
numerical computations and generating the intensity plot. The remaining ones
are needed to store the plots as image files.
\begin{verbatim}
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.patches as patches

import glob
from PIL import Image
import re
import os
\end{verbatim}

We define the physical constants used in the problem.
\begin{verbatim}
c = 3e8           # The speed of light in vacuum.
epsr = 1e4        # The relative dielectric permittivity of the screen.
sigma = 0.3       # The conductivity of the screen.
epsz = 8.854e-12  # Permittivity of free space in SI units.
\end{verbatim}

The geometrical parameters of the problem are:
\begin{verbatim}
# The extent of the domain. **You can alter these parameters.**
ie = 200 # vertical limit.
je = 200 # horizontal limit.

# The central points in either directions.
ic = ie // 2 - 1
jc = je // 2 - 1

# The location of the perfectly matching layer.
ia = 10           # starting vertical position after PML.
ib = ie - ia - 1  # ending horizontal position before PML.
ja = 10           # starting horizontal position after PML.
jb = je - ja - 1  # ending horizontal position before PML.

# Slit geometry. **You can alter these parameters.**
slit_width = 8 # width of each slit (vertical measure).
slit_sep = 20 # separation between both slits (vertical measure).
screen_width = 8 # width of the screen (horizontal measure).
\end{verbatim}

We now allocate space to store the two dimensional field vectors. We simulate
a TM wave propagating in vacuum until it hits the screen with two slits in 
it.
\begin{verbatim}
ez = np.zeros((ie, je)) # electric field.
dz = np.zeros((ie, je)) # electric displacement.
hx = np.zeros((ie, je)) # x-component of H field.
hy = np.zeros((ie, je)) # y-component of H field.
iz = np.zeros((ie, je))
ihx = np.zeros((ie, je))
ihy = np.zeros((ie, je))

ez_inc = np.zeros(je) # incident electric field
hx_inc = np.zeros(je) # x-component of incident H field
\end{verbatim}

The simulation parameters and certain auxiliary variables are defined in the
code snippet below.
\begin{verbatim}
ddx = 0.01 # Cell size in m
dt = ddx / 2*c
gaz = np.ones((ie, je))
gbz = np.zeros((ie, je))

boundary_low = [0, 0]
boundary_high = [0, 0]
# Time for which the source is generated. 
# **You can alter this parameter.**
t0 = 10       
# Number of time steps. 
# **You can alter this parameter.**
nsteps = 500

# Vectors used for PML.
gi2 = np.ones(ie)
gi3 = np.ones(ie)
fi1 = np.zeros(ie)
fi2 = np.ones(ie)
fi3 = np.ones(ie)

gj2 = np.ones(je)
gj3 = np.ones(je)
fj1 = np.zeros(je)
fj2 = np.ones(je)
fj3 = np.ones(je)

# Dictionary to keep track of desired points for plotting.
plotting_points = []
for num_steps in range(1,nsteps+1):
    plotting_points.append({'label': 'a',
                            'num_steps': num_steps,
                            'data_to_plot': None})
\end{verbatim}

We now define the functions that do the FDTD computations. The first one
creates the dielectric profile of the domain. It initializes the portion of the
domain that has the screen with the screens $\sigma$ and $\kappa$.
\begin{verbatim}
def build_dielectric_profile():
  for j in range(ja, jb): # vertical range
      for i in range(ia, ib): # horizontal range
          if (
              ((i > 0 and i < ic - slit_width/2-slit_sep/2) or
              (i > ic - slit_sep/2 + slit_width/2 and 
			   i < ic + slit_sep/2 - slit_width/2) or
              (i > ic + slit_width/2+slit_sep/2 and i < ie))
              and (j > jc-screen_width/2 and j < jc+screen_width/2)
              ): # If position is inside the dielectric

              gaz[i, j] = 1 / (epsr + (sigma * dt / epsz))
              gbz[i, j] = (sigma * dt / epsz)
\end{verbatim}

The next function implements the absorbing boundary condition with a perfectly
matched layer.
\begin{verbatim}
def create_pml():
  npml = ia + 1 # number of points for the PML
  for n in range(npml):
      xnum = npml - n
      xd = npml
      xxn = xnum / xd
      xn = 0.33 * xxn ** 3

      gi2[n] = 1 / (1 + xn)
      gi2[ie - 1 - n] = 1 / (1 + xn)
      gi3[n] = (1 - xn) / (1 + xn)
      gi3[ie - 1 - n] = (1 - xn) / (1 + xn)

      gj2[n] = 1 / (1 + xn)
      gj2[je - 1 - n] = 1 / (1 + xn)
      gj3[n] = (1 - xn) / (1 + xn)
      gj3[je - 1 - n] = (1 - xn) / (1 + xn)

      xxn = (xnum - 0.5) / xd
      xn = 0.33 * xxn ** 3

      fi1[n] = xn
      fi1[ie - 2 - n] = xn
      fi2[n] = 1 / (1 + xn)
      fi2[ie - 2 - n] = 1 / (1 + xn)
      fi3[n] = (1 - xn) / (1 + xn)
      fi3[ie - 2 - n] = (1 - xn) / (1 + xn)

      fj1[n] = xn
      fj1[je - 2 - n] = xn
      fj2[n] = 1 / (1 + xn)
      fj2[je - 2 - n] = 1 / (1 + xn)
      fj3[n] = (1 - xn) / (1 + xn)
      fj3[je - 2 - n] = (1 - xn) / (1 + xn)
\end{verbatim}

The fields are computed in
\begin{verbatim}
def compute_fields(time_step):
  # Print progress.
  if time_step%10 == 0:
      print("\r",100*time_step/nsteps,'%', end="")

  for j in range(1, je):
      ez_inc[j] = ez_inc[j] + 0.5 * (hx_inc[j - 1] - hx_inc[j])

  # Absorbing Boundary Conditions
  ez_inc[0] = boundary_low.pop(0)
  boundary_low.append(ez_inc[1])
  ez_inc[je - 1] = boundary_high.pop(0)
  boundary_high.append(ez_inc[je - 2])

  for j in range(1, je):
      for i in range(1, ie):
          dz[i, j] = gi3[i] * gj3[j] * dz[i, j] + \
                      gi2[i] * gj2[j] * 0.5 * \
                      (hy[i, j] - hy[i - 1, j] -
                      hx[i, j] + hx[i, j - 1])

  # Source of electromagnetic waves.
  ez_inc[1] = (np.sin(2*np.pi*time_step/(2*t0)))**2

  for i in range(ia, ib + 1):
      dz[i, ja] = dz[i, ja] + 0.5 * hx_inc[ja - 1]
      dz[i, jb] = dz[i, jb] - 0.5 * hx_inc[jb]

  for j in range(0, je):
      for i in range(0, ie):
          ez[i, j] = gaz[i, j] * (dz[i, j] - iz[i, j])
          iz[i, j] = iz[i, j] + gbz[i, j] * ez[i, j]

  for j in range(0, je - 1):
      hx_inc[j] = hx_inc[j] + 0.5 * (ez_inc[j] - ez_inc[j + 1])

  for j in range(0, je - 1):
      for i in range(0, ie - 1):
          curl_e = ez[i, j] - ez[i, j + 1]
          ihx[i, j] = ihx[i, j] + curl_e
          hx[i, j] = fj3[j] * hx[i, j] + fj2[j] * \
                      (0.5 * curl_e + fi1[i] * ihx[i, j])

  for i in range(ia, ib + 1):
      hx[i, ja - 1] = hx[i, ja - 1] + 0.5 * ez_inc[ja]
      hx[i, jb] = hx[i, jb] - 0.5 * ez_inc[jb]

  for j in range(0, je):
      for i in range(0, ie - 1):
          curl_e = ez[i, j] - ez[i + 1, j]
          ihy[i, j] = ihy[i, j] + curl_e
          hy[i, j] = fi3[i] * hy[i, j] - fi2[i] * \
                      (0.5 * curl_e + fj1[j] * ihy[i, j])

  for j in range(ja, jb + 1):
      hy[ia - 1, j] = hy[ia - 1, j] - 0.5 * ez_inc[j]
      hy[ib, j] = hy[ib, j] + 0.5 * ez_inc[j]

  # Save data at certain points for later plotting
  for plotting_point in plotting_points:
      if time_step == plotting_point['num_steps']:
          plotting_point['data_to_plot'] = np.copy(ez)
\end{verbatim}

The main simulation is carried out by calling the functions in the previous
snippets.
\begin{verbatim}
build_dielectric_profile()
create_pml()

print('Starting FDTD...')
for time_step in range(1, nsteps + 1):
  compute_fields(time_step)
\end{verbatim}

The final snippet creates the intensity profile and saves intermediate results
in image files.
\begin{verbatim}
plt.rcParams['font.size'] = 12
plt.rcParams['grid.color'] = 'gray'
plt.rcParams['grid.linestyle'] = 'dotted'

print('Starting to save images to make gif...')
for subplot_num, plotting_point in enumerate(plotting_points):
    if subplot_num%10 == 0:
        print("\r",100*subplot_num/nsteps,'%', end="")

    fig = plt.figure()
    ax = fig.add_subplot(1, 1, 1)

    # Plot the dielectric in white color
    rect1 = patches.Rectangle((jc-screen_width/2,0),
                              screen_width,
                              ic-slit_sep/2-slit_width/2,
                              linewidth=1,
                              edgecolor='#ffffff',
                              facecolor='#ffffff')
    rect2 = patches.Rectangle((jc-screen_width/2,
                               ic-slit_sep/2+slit_width/2),
                              screen_width,
                              slit_sep-slit_width,
                              linewidth=1,
                              edgecolor='#ffffff',
                              facecolor='#ffffff')
    rect3 = patches.Rectangle((jc-screen_width/2,
                               ic+slit_sep/2+slit_width/2),
                              screen_width,
                              ic-slit_sep/2-slit_width/2,
                              linewidth=1,
                              edgecolor='#ffffff',
                              facecolor='#ffffff')
    ax.add_patch(rect1)
    ax.add_patch(rect2)
    ax.add_patch(rect3)

    plt.imshow(plotting_point['data_to_plot'],
               interpolation='bilinear',
               cmap='gist_heat')
    plt.savefig('figs_%s.png'%str(subplot_num+1))
    plt.close();

def tryint(s):
    try:
        return int(s)
    except:
        return s

def alphanum_key(s):
    """ Turn a string into a list of string and number chunks.
        "z23a" -> ["z", 23, "a"]
    """
    return [ tryint(c) for c in re.split('([0-9]+)', s) ]

def sort_nicely(l):
    """ Sort the given list in the way that humans expect.
    """
    l.sort(key=alphanum_key)
    return l

# filepaths
if not os.path.exists('figs'):
    print('Folder "figs" already exists. Saving/replacing images inside it...')
    os.makedirs('figs')
else:
    print('Creating folder "figs". Saving images inside it...')

fp_in = "figs_*.png"
fp_out = "YoungMovie.gif"

img, *imgs = [Image.open(f) for f in sort_nicely(glob.glob(fp_in))]
# duration: time of each frame in miliseconds
img.save(fp=fp_out, format='GIF', append_images=imgs,
         save_all=True, duration=20, loop=0, optimize=True)

x = []
x1 = 40 # Starting vertical position
x2 = 160 # Final vertical position

# Construct array with x values, along vertical position from x1 to x2
for i in range(x1,x2):
    x.append(i)

# Here we compute an averaged field over two times t1 and t2
average = 0
t1 = nsteps - 100
t2 = nsteps - 1
y_measure = 150 # Horizontal position where the filed is to be measured

for i in range(t1,t2):
    average += abs(plotting_points[i]['data_to_plot'][x1:x2,y_measure])

# Plot the averaged field
ignore = plt.plot(x, average/(t2-t1))
\end{verbatim}

The intensity profile generated with $500$ steps of simulation are 
shown in figure \ref{f4}. Figure \ref{f5} shows the interference 
pattern.
\begin{figure}
\centering
\includegraphics[scale=0.5]{fdtd-3}
\caption{Intensity profile in Young's experiment}\label{f4}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=0.5]{figs_499}
\caption{Interference pattern in Young's experiment}\label{f5}
\end{figure}

\bibliographystyle{plain}
\bibliography{em}
\end{document}
